Data Analysis Extensions

### 2016-02-06 RFCs/HDF5_Library/AnalysisExt/analysis_ext-body.tex
More update to analysis document
Former-commit-id: bc183e6cb18f23e94a8800a6dbc554da7d4c95b2

\section{Index Objects}
Index objects are designed to accelerate creation of views from query operations.
For example, if the previously described ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$'' query
is going to be either frequently executed or executed
on a container that contains significantly large datasets, indexes
could be created in that container, which would speed up the
creation of views when querying for link names and for data element values.
Indexes created for accelerating the  ``$\mathrm{(link\_name = Pressure)}$'' or
``$\mathrm{(data\_element}$ $\mathrm{= 17)}$'' queries
would also improve view creation for the more complex
``$\mathrm{(link\_name = Pressure)}$ $\mathrm{\land}$
$\mathrm{(data\_element = 17)}$'' query.
Although creating indexes for metadata components of queries, such as link or
attribute names, is possible, we first focus on index creation for dataset elements,
as they represent the largest volume of data in typical HPC application usage of
HDF5. Queries with metadata components execute properly,
but are not able to be accelerated with an index currently.

The indexing API can work in conjunction with the view creation. When an \texttt{H5Qapply}
call is made on a given location, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.
There are different techniques for creating data element indexes, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

\subsection{Indexing Interface and Plugins}

@@

    /* Callbacks, described above */
    void *(*create)(hid_t dataset_id, hid_t xcpl_id, hid_t xapl_id,
        size_t *metadata_size, void **metadata);
    herr_t (*remove)(hid_t dataset_id, size_t metadata_size, void *metadata);
    void *(*open)(hid_t dataset_id, hid_t xapl_id, size_t metadata_size,
        void *metadata);
    herr_t (*close)(void *idx_handle);
    herr_t (*copy)(hid_t src_dataset_id, hid_t dest_dataset_id, hid_t xcpl_id,
        hid_t xapl_id, size_t src_metadata_size, void *src_metadata,
        size_t *dest_metadata_size, void **dest_metadata);
    herr_t (*pre_update)(void *idx_handle, hid_t dataspace_id, hid_t xxpl_id);
    herr_t (*post_update)(void *idx_handle, const void *buf, hid_t dataspace_id,
        hid_t xxpl_id);
    herr_t (*query)(void *idx_handle, hid_t query_id, hid_t xxpl_id,
        hid_t *dataspace_id);
    herr_t (*refresh)(void *idx_handle, size_t *metadata_size, void **metadata);
    herr_t (*get_size)(void *idx_handle, hsize_t *idx_size);
} H5X_class_t;
\end{lstlisting}
}

@@

\subsection{Limitations}

@@

\subsubsection*{Questions}
Some questions are still open regarding the handling of indexes:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
\textit{In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here.
We could mark the index as out of date and rebuild the index when the plugin is
available again.}
\item How to handle index queries when the specified index plugin is not
available? \textit{We could fallback to another plugin and do a brute force query on the data.}
\end{enumerate}

\subsection{Support for HDF5 Compound Types (TBD)}

@@

\subsection{Support for HDF5 Chunking (TBD)}

@@

\subsection{Support for Parallel Indexing (TBD)}

@@

>>>>

\section{Index Objects}
Index objects are designed to accelerate creation of views from query operations.
For example, if the previously described ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$'' query
is going to be either frequently executed or executed
on a container that contains significantly large datasets, indexes
could be created in that container, which would speed up the
creation of views when querying for link names and for data element values.
Indexes created for accelerating the  ``$\mathrm{(link\_name = Pressure)}$'' or
``$\mathrm{(data\_element}$ $\mathrm{= 17)}$'' queries
would also improve view creation for the more complex
``$\mathrm{(link\_name = Pressure)}$ $\mathrm{\land}$
$\mathrm{(data\_element = 17)}$'' query.
We distinguish two different types of indexes: \textit{data} indexes and
\textit{metadata} indexes. Data indexes apply to dataset elements, which
represent the largest volume of data in typical HPC application usage of HDF5,
whereas metadata indexes apply to link or attribute name components of the queries.

\subsection{Data Indexing}

The data indexing API can work in conjunction with the view creation. When an \texttt{H5Qapply}
call is made on a given location, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.
There are different techniques for creating data element indexes, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

\subsubsection{Indexing Interface and Plugins}

@@

    /* Callbacks */
    union {               /* Union of callback index structures */
        H5X_data_class_t data_class;
        H5X_metadata_class_t metadata_class;
    } idx_class;
} H5X_class_t;

typedef struct {
    void *(*create)(hid_t dataset_id, hid_t xcpl_id, hid_t xapl_id,
        size_t *metadata_size, void **metadata);
    herr_t (*remove)(hid_t dataset_id, size_t metadata_size, void *metadata);
    void *(*open)(hid_t dataset_id, hid_t xapl_id, size_t metadata_size,
        void *metadata);
    herr_t (*close)(void *idx_handle);
    herr_t (*copy)(hid_t src_dataset_id, hid_t dest_dataset_id, hid_t xcpl_id,
        hid_t xapl_id, size_t src_metadata_size, void *src_metadata,
        size_t *dest_metadata_size, void **dest_metadata);
    herr_t (*pre_update)(void *idx_handle, hid_t dataspace_id, hid_t xxpl_id);
    herr_t (*post_update)(void *idx_handle, const void *buf, hid_t dataspace_id,
        hid_t xxpl_id);
    herr_t (*query)(void *idx_handle, hid_t query_id, hid_t xxpl_id,
        hid_t *dataspace_id);
    herr_t (*refresh)(void *idx_handle, size_t *metadata_size, void **metadata);
    herr_t (*get_size)(void *idx_handle, hsize_t *idx_size);
} H5X_data_class_t;
\end{lstlisting}
}

@@

\subsubsection{Limitations}

@@

\paragraph{Questions}Some questions are still open regarding the handling of indexes:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
\textit{In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here.
We could mark the index as out of date and rebuild the index when the plugin is
available again.}
\item How to handle index queries when the specified index plugin is not
available? \textit{We could fallback to another plugin and do a brute force query on the data.}
\end{enumerate}

\subsubsection{Support for HDF5 Compound Types (TBD)}

@@

\subsubsection{Support for HDF5 Chunking (TBD)}

@@

\subsubsection{Support for Parallel Indexing (TBD)}

@@

\subsection{Metadata Indexing}

Metadata indexing accelerates HDF5 metadata query operations, improving 
the speed of data analysis operations. Currently metadata query operations
in HDF5 are performed by traversing the contents of the container specified
in the query, potentially generating a large amount of I/O and taking
significant time to create the query results. Adding a mechanism for storing
metadata indexes to HDF5 containers can accelerate query operations by avoiding
these expensive traversals, as is available for data element queries.

A metadata index is a data structure that tracks a particular aspect of an HDF5 container 
and can quickly answer queries about that aspect. For example, a metadata index can 
be created that tracks all the link names for objects in a container, updating the index 
whenever a link is added, modified or removed. This index can then be applied to 
queries on link names, avoiding a traversal over all the links for groups in the container.

\subsubsection{Indexing Interface and Plugins}

The interface for metadata indexes in HDF5 containers is modeled after the interface 
for data element indexes in HDF5, with a plugin architecture that allows new metadata 
index packages to be developed and added at application runtime, without modifications 
to the core HDF5 library.
The HDF5 library can then interact with the third-party indexes created in this manner 
through a set of callback interfaces that present query information to the
indexing package and retrieve results from it.
An HDF5 metadata index can be considered as a generic key-value store, in the
sense that for each \textit{key} that represents an HDF5 object ID, is associated
either a link/attribute name or an attribute \textit{value}.

{\lstsetc
\begin{lstlisting}
typedef struct {
    void *(*create)(hid_t loc_id, hid_t xcpl_id, hid_t xapl_id,
        size_t *metadata_size, void **metadata);
    herr_t (*remove)(hid_t loc_id, size_t metadata_size, void *metadata);
    void *(*open)(hid_t loc_id, hid_t xapl_id, size_t metadata_size,
        void *metadata);
    herr_t (*close)(void *idx_handle);

    herr_t (*insert_entry)(void *idx_handle, hid_t obj_id, hid_t xxpl_id);
    herr_t (*remove_entry)(void *idx_handle, hid_t obj_id, hid_t xxpl_id);
    herr_t (*query)(void *idx_handle, hid_t query_id, hid_t xxpl_id,
        size_t *obj_count, hid_t *obj_ids[]);

    herr_t (*get_size)(void *idx_handle, hsize_t *idx_size);
} H5X_metadata_class_t;
\end{lstlisting}
}

The \texttt{create}, \texttt{remove}, \texttt{open}, \texttt{close} callbacks
are similar to the data index plugins, albeit they operate on a file location and
not on a single object. The \texttt{insert\_entry}, \texttt{remove\_entry} callbacks
respectively insert/remove a referenced object from the index. Values can be extracted from
the object ID, depending on the metadata type that the plugin will index.
When the \texttt{query} callback is invoked, a list of object IDs that matches
the query is returned. The \texttt{get\_size} callback returns the size of the index
and additional info callbacks may also be added depending on the needs.

Are considered both a MDHIM and a B-Tree plugin.
